---
title: "functions.R"
author: "SVD"
date: "7 April 2016"
output:
  rmdformats::readthedown:
    highlight: kate
    number_sections: yes
    code_folding: show
---


\newpage

I have attached here some functions that are great for working in R. These functions have been obtained from the web, friends, teachers and colleagues working and analyzing in R.

#  To filter a dataframe to remove rows containing NA for a specific variable `column`:


```{r}
library(tidyverse)
library(magrittr)
library(broom)
library(plyr)
```

```{r}

completeFun <- function(data, desiredCols) {
  completeVec <- complete.cases(data[, desiredCols])
  return(data[completeVec, ])
}

# data = dataframe
# desiredCols = name of column/s that we want to remove NA from 
```

# This function will help to re-arrange columns in a dataframe: 

```{r}
moveme <- function (invec, movecommand) {
  movecommand <- lapply(strsplit(strsplit(movecommand, ";")[[1]], 
                                 ",|\\s+"), function(x) x[x != ""])
  movelist <- lapply(movecommand, function(x) {
    Where <- x[which(x %in% c("before", "after", "first", 
                              "last")):length(x)]
    ToMove <- setdiff(x, Where)
    list(ToMove, Where)
  })
  myVec <- invec
  for (i in seq_along(movelist)) {
    temp <- setdiff(myVec, movelist[[i]][[1]])
    A <- movelist[[i]][[2]][1]
    if (A %in% c("before", "after")) {
      ba <- movelist[[i]][[2]][2]
      if (A == "before") {
        after <- match(ba, temp) - 1
      }
      else if (A == "after") {
        after <- match(ba, temp)
      }
    }
    else if (A == "first") {
      after <- 0
    }
    else if (A == "last") {
      after <- length(myVec)
    }
    myVec <- append(temp, values = movelist[[i]][[1]], after = after)
  }
  myVec
}

# this is primarily for vectors 
# eg. moveMe(names(df),"g first") = will insert g as the first column
# moveMe(names(df), "g first; a last; e before c") = will do exactly as written in the code
# for dataframes , we need to use: df[moveMe(names(df),"g first")] 
```


# Will create a series of histograms for a dataframe containing only continuous variables:

```{r}
plotHistFunc <- function(x, na.rm = TRUE, ...) {
  nm <- names(x)
  for (i in seq_along(nm)) {
    print(ggplot(x,aes_string(x = nm[i])) + geom_histogram(alpha = .5,fill = "mediumseagreen",col="darkblue")) }
}


# 1. using select in dplyr, can create a dataframe which will contain only continuous variables
#2. Then with this function plotHistFunc(dataframe, na.rm=TRUE) we can create a series of histograms of all the variables in the dataframe.
```

# Two x two function ; Prof. Love: 

```{r}

`twobytwo` <-
  function(a,b,c,d, namer1 = "Row1", namer2 = "Row2", namec1 = "Col1", namec2 = "Col2")
    # build 2 by 2 table and run Epi library's twoby2 command to summarize
    # from the row-by-row counts in a cross-tab
    # upper left cell is a, upper right is b, lower left is c, lower right is d
    # names are then given in order down the rows then across the columns
    # use standard epidemiological format - outcomes in columns, treatments in rows
  {
    require(Epi)
    .Table <- matrix(c(a, b, c, d), 2, 2, byrow=T, dimnames=list(c(namer1, namer2), c(namec1, namec2)))
    twoby2(.Table)
  }

# as described will create a two by two table with OR,RR and RD as results
```


# This will create a histogram, boxplot and standard summary statistics for a variable; Prof.Love: 

```{r}
`eda6` <-
  function(a.raw)
  {
    a <- na.omit(a.raw)
    
    if( is.null(names(a.raw) ) ){ mname = substitute(a.raw) }
    else{ mname = names(a.raw) }
    
    ## Histogram and Normal Q-Q Plot  
    par(mfrow=c(1,2))
    hist(a, prob=T, main=paste("Histogram & Normal Fit for", mname), col="wheat", xlab=paste(mname), ylab="")
    curve(dnorm(x, mean=mean(a), sd = sd(a)), col="red", lty=1, lwd=1.5, add=TRUE)
    library(car)
    qqPlot(a, main=paste("Normal Q-Q Plot for", mname), 
           xlab="Standard Normal Quantiles", 
           ylab=paste("Sorted ", mname, " values"), cex=1.5)
    ## this version of a Normal Q-Q plot, including pointwise confidence intervals
    par(mfrow=c(1,1))
    
    ## Boxplot    
    boxplot(a, cex=1.5, horizontal=TRUE, main=paste("Boxplot and Mean +/- SD of", mname), xlab=paste(mname), col="wheat")
    points(mean(a), 1.3, col="blue", pch=16, cex=1.5)
    arrows(mean(a)-sd(a), 1.3, mean(a)+sd(a), 1.3, 
           code=3, col="blue", angle=40, length=.1, lwd=1.5)
    
    ## Favorite Summary Statistics using mosaic library
    library(mosaic, quietly=TRUE, warn.conflicts=FALSE) 
    
    b <- list("Variable" = mname, 
              "Favorite Descriptive Statistics" = favstats(a), 
              "Most Common Value(s)" = names(table(a))[table(a)==max(table(a))],
              "Nonparametric Skew" = round((mean(a) - median(a))/sd(a),2))
    return(b)
    
    rm(a, b)
  }


# eda6(df$var)
```

# Bootstrapped difference in means:

```{r}
`bootdif` <-
  function(y, g, conf.level=0.95, B.reps = 2000) {
    require(Hmisc)
    lowq = (1 - conf.level)/2
    g <- as.factor(g)
    a <- attr(smean.cl.boot(y[g==levels(g)[1]], B=B.reps, reps=TRUE),'reps')
    b <- attr(smean.cl.boot(y[g==levels(g)[2]], B=B.reps, reps=TRUE),'reps')
    meandif <- diff(tapply(y, g, mean, na.rm=TRUE))
    a.b <- quantile(b-a, c(lowq,1-lowq))
    res <- c(meandif, a.b)
    names(res) <- c('Mean Difference',lowq, 1-lowq)
    res
  }

# here y = continuous variable, g = grouping variable
```

# Obtaining correlation plot matrix for starting linear regression, Prof.Love:

```{r}
### panel.hist function from Chang R Graphics Cookbook

`panel.hist` <- function(x, ...)
{
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(usr[1:2], 0, 1.5) )
  h <- hist(x, plot = FALSE)
  breaks <- h$breaks; nB <- length(breaks)
  y <- h$counts; y <- y/max(y)
  rect(breaks[-nB], 0, breaks[-1], y, col = "cyan", ...)
}

### panel.cor function from Chang R Graphics Cookbook
`panel.cor` <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- cor(x, y)
  txt <- format(c(r, 0.123456789), digits = digits)[1]
  txt <- paste0(prefix, txt)
  if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
  text(0.5, 0.5, txt, cex = cex.cor * (1 + abs(r)) / 2)
}

#pairs (~ sbp + age + weight + height, data=wcgs,main="Augmented Scatterplot Matrix",upper.panel = panel.smooth,diag.panel = panel.hist,lower.panel =panel.cor)

# using the above code and inserting the variables as shown, we can get a scatterplot matrix : T.Love and Winston Chang.

```


# Code to create a `Love-plot` in ggplot2, Dr Ahmad Al-khalil : 




Building Love Plot
covnames <- covlist
pre.szd <- NULL; post.szd <- NULL
for(i in 1:length(covnames)) {
  pre.szd[i] <- mb1$ BeforeMatching[[i]]$sdiff.pooled
  post.szd[i] <- mb1$ AfterMatching[[i]]$sdiff.pooled
}

temp <- data.frame(pre.szd, post.szd, row.names=covnames)


temp1 <- data.frame(pre.szd, post.szd, covnames)#combine the 3 variable into a data.base.
library(dplyr)
temp1<-temp1 %>% arrange(pre.szd)%>% mutate (covnames= factor(covnames,covnames)) #use dplyr mutate and arrange function to sort the covariates by pre.szd.
library(reshape2)
temp2<-melt(temp1,measure.vars=c("pre.szd","post.szd")) ##melt the database base to get a two level variable ("pre.szd","post.szd") that I will use in the plot.

p<-ggplot(temp2, aes(value, covnames, shape= variable,colour=variable)) ## the first step is tell ggplot that I want to use temp2 database to plot the (value) variable on x axis, covnames on y axis. Then I tell it to differentiate between the plotted geom based on (variable)  which is two level categorical variable ("pre.szd","post.szd").

p<-p +geom_point(size =3)+theme_bw()+labs(title="Standardized Difference Plot")+labs(x="Standardized Difference (%)")+labs(y="Covariates") # here, I tell ggplot that I want the values to be plotted as point (geom_point), I chose size 3, I change the background theme to black and white, then I label the axes.

p<-p+theme(legend.title=element_blank())+scale_colour_manual(values = c("pre.szd"="red", "post.szd"="green"),labels=c("Before Matching","After Matching"))+scale_shape_discrete(labels=c("Before Matching","After Matching")) ## here, I drop the legend title, I assign specific colors to the points, I rename the legend's labels.

p<-p + theme(legend.justification=c(1,0), legend.position=c(1,0)) + geom_vline(xintercept = 0,colour="blue") + geom_vline(xintercept = c(-10,10) ,colour="blue",linetype = "longdash") # I reposition the legend to right lower corner, I add the vertical lines.


p+ coord_cartesian(xlim = c(-60, 50)) + scale_x_continuous(breaks = seq(from = -200, to = 200, by = 10)) # Finally, I change x-axis limits and change the breaks to every 10.




# Pirate plot: 

`yarrr` library: Use the pirate plot to present the RDI trifecta of data : 1. Raw data = presents all datapoints 2. Presents mean and central distribution 3. I = inferential statistics ; presents High Density Intervals using Bayesian estimates supersedes T-tests method.

# Presents the confidence interval with the `SAIFS` method:

```{r}
`saifs.ci` <- 
  function(x, n, conf.level=0.95, dig=3)
  {
    p.sample <- round(x/n, digits=dig)
    
    p1 <- x / (n+1)
    p2 <- (x+1) / (n+1)
    
    var1 <- (p1*(1-p1))/n
    se1 <- sqrt(var1)
    var2 <- (p2*(1-p2))/n
    se2 <- sqrt(var2)
    
    lowq = (1 - conf.level)/2
    tcut <- qt(lowq, df=n-1, lower.tail=FALSE)
    
    lower.bound <- round(p1 - tcut*se1, digits=dig)
    upper.bound <- round(p2 + tcut*se2, digits=dig)
    res <- c(p.sample, lower.bound, upper.bound)
    names(res) <- c('Sample Proportion',lowq, 1-lowq)
    res
  }

```

# RMSE and MAE: 

Functions created to obtain `root mean square error` and `mean absolute error` from linear models.

```{r}

rmse <- function(actual,predicted)
{
  sqrt(mean((actual- predicted)^2))
}

mae <- function(actual,predicted)
{
  mean(abs(actual-predicted))
}

```

# `ggsurv` function:

For creating KM curves in ggplot2. Also can use `autoplot` from the the`survMisc` library. When using `ggsurv` for KM curves, use geom_ribbon to create CI bands for the groups.

```{r}
#geom_ribbon(aes(ymin = low, ymax = up, fill=group),alpha = 0.3) + ggplot2::scale_fill_discrete(guide=FALSE) ## and then use the following code to remove the legend from the plot, so that we dont have two legends.
```


# Creating series of box-plots with ggplot2:

```{r}

# This function will create sequential box-plots for all continuous variables in a dataframe. 
# Initially create a dataframe with the categorical variable as the first column and all continuous variables after that. This can be created using the help of `moveme` function listed earlier and the `dplyr` library.
# To use the dataframe, simply type plotBoxfunc(df).
# The function will automatically remove those rows with missing data in each continuous variable.
# The ggplot2 library needs to be called for the function to work. Please do that initially before running function.
# The function will print the boxplots one below the other with legend at the right side as in the normal ggplot2 method.
# fill color is set to categorical variable. 

plotBoxfunc <- function(df,na.rm=TRUE,...){ 
  
  x = names(df[1,])
  nm = names(df[,2:ncol(df)])
  for (i in seq_along(nm)){
    print(ggplot(df,aes_string(x = x, y = nm[i],fill=x))+ geom_boxplot() + theme_bw())
  }
}
```


# Creating publication ready graphs with ggplot2

Use the `cowplot` library to create publication ready graphs with `ggplot2`. The details are available at the URL : https://cran.r-project.org/web/packages/cowplot/vignettes/introduction.html
The `cowplot` library will create graphs which will not have any background at all. 



# Changing labels in the legend:

```{r}

# this code will allow change in the legend labels using the code with the `fill` code (1) with this we can change the labels of the categorical labels (2) using `scale_fill_manual` code, we can change the heading of the legend as well as the color of the two density curves and also the colors in the legend.


#p <- ggplot(p6,aes(x=ps, fill= factor(weight, labels = c("Normal","Morbidly obese")))) + geom_density(alpha=0.25)+ xlab("Propensity scores") + ylab("Density function") + ggtitle("Density plot of propensity scores in both cohorts") + scale_fill_manual(name = "Cohort", values = c("red","blue")) + theme_bw()

```


# Adding CI as ribbon in `ggsurv` function with only one legend:

```{r}
ggsurv.svd <- function(s, CI = 'def', plot.cens = T, surv.col = 'gg.def',
                   cens.col = 'red', lty.est = 1, lty.ci = 2,
                   cens.shape = 3, back.white = F, xlab = 'Time',
                   ylab = 'Survival', main = ''){
  
  library(ggplot2)  
  strata <- ifelse(is.null(s$strata) ==T, 1, length(s$strata))
  stopifnot(length(surv.col) == 1 | length(surv.col) == strata)
  stopifnot(length(lty.est) == 1 | length(lty.est) == strata)
  
  ggsurv.s <- function(s, CI = 'def', plot.cens = T, surv.col = 'gg.def',
                       cens.col = 'red', lty.est = 1, lty.ci = 2,
                       cens.shape = 3, back.white = F, xlab = 'Time',
                       ylab = 'Survival', main = ''){
    
    dat <- data.frame(time = c(0, s$time),
                      surv = c(1, s$surv),
                      up = c(1, s$upper),
                      low = c(1, s$lower),
                      cens = c(0, s$n.censor))
    dat.cens <- subset(dat, cens != 0)
    
    col <- ifelse(surv.col == 'gg.def', 'black', surv.col)
    
    pl <- ggplot(dat, aes(x = time, y = surv)) + 
      xlab(xlab) + ylab(ylab) + ggtitle(main) + 
      geom_step(col = col, lty = lty.est) + scale_fill_discrete(guide = FALSE)
    
    pl <- if(CI == T | CI == 'def') {
      pl + geom_ribbon(aes(ymin = low, ymax = up, fill=group),alpha = 0.3)
    } else (pl)
    
    pl <- if(plot.cens == T & length(dat.cens) > 0){
      pl + geom_point(data = dat.cens, aes(y = surv), shape = cens.shape,
                      col = cens.col)
    } else if (plot.cens == T & length(dat.cens) == 0){
      stop ('There are no censored observations') 
    } else(pl)
    
    pl <- if(back.white == T) {pl + theme_bw()
    } else (pl)
    pl
  }
  
  ggsurv.m <- function(s, CI = 'def', plot.cens = T, surv.col = 'gg.def',
                       cens.col = 'red', lty.est = 1, lty.ci = 2,
                       cens.shape = 3, back.white = F, xlab = 'Time',
                       ylab = 'Survival', main = '') {
    n <- s$strata
    
    groups <- factor(unlist(strsplit(names
                                     (s$strata), '='))[seq(2, 2*strata, by = 2)])
    gr.name <-  unlist(strsplit(names(s$strata), '='))[1]
    gr.df <- vector('list', strata)
    ind <- vector('list', strata)
    n.ind <- c(0,n); n.ind <- cumsum(n.ind)
    for(i in 1:strata) ind[[i]] <- (n.ind[i]+1):n.ind[i+1]
    
    for(i in 1:strata){
      gr.df[[i]] <- data.frame(
        time = c(0, s$time[ ind[[i]] ]),
        surv = c(1, s$surv[ ind[[i]] ]),
        up = c(1, s$upper[ ind[[i]] ]), 
        low = c(1, s$lower[ ind[[i]] ]),
        cens = c(0, s$n.censor[ ind[[i]] ]),
        group = rep(groups[i], n[i] + 1)) 
    }
    
    dat <- do.call(rbind, gr.df)
    dat.cens <- subset(dat, cens != 0)
    
    pl <- ggplot(dat, aes(x = time, y = surv, group = group)) + 
      xlab(xlab) + ylab(ylab) + ggtitle(main) + 
      geom_step(aes(col = group, lty = group)) + scale_fill_discrete(guide = FALSE)
    
   #col <- if(length(surv.col == 1)){
    #  scale_colour_manual(name = gr.name, values = rep(surv.col, strata))
    #} else{
     # scale_colour_manual(name = gr.name, values = surv.col)
    #}
    
    pl <- if(surv.col[1] != 'gg.def'){
      pl + col
    } else {pl + scale_colour_discrete(name = gr.name)}
    
    line <- if(length(lty.est) == 1){
      scale_linetype_manual(name = gr.name, values = rep(lty.est, strata))
    } else {scale_linetype_manual(name = gr.name, values = lty.est)}
    
    pl <- pl + line
    
    pl <- if(CI == T) {
      if(length(surv.col) > 1 && length(lty.est) > 1){
        stop('Either surv.col or lty.est should be of length 1 in order
             to plot 95% CI with multiple strata')
      }else if((length(surv.col) > 1 | surv.col == 'gg.def')[1]){
        pl + geom_ribbon(aes(ymin = low, ymax = up, fill=group),alpha = 0.3)
      } else{pl +  geom_step(aes(y = up, lty = group), col = surv.col) +
          geom_step(aes(y = low,lty = group), col = surv.col)}   
    } else {pl}
    
    
    pl <- if(plot.cens == T & length(dat.cens) > 0){
      pl + geom_point(data = dat.cens, aes(y = surv), shape = cens.shape,
                      col = cens.col)
    } else if (plot.cens == T & length(dat.cens) == 0){
      stop ('There are no censored observations') 
    } else(pl)
    
    pl <- if(back.white == T) {pl + theme_bw()
    } else (pl) 
    pl
  } 
  pl <- if(strata == 1) {ggsurv.s(s, CI , plot.cens, surv.col ,
                                  cens.col, lty.est, lty.ci,
                                  cens.shape, back.white, xlab,
                                  ylab, main) 
  } else {ggsurv.m(s, CI, plot.cens, surv.col ,
                   cens.col, lty.est, lty.ci,
                   cens.shape, back.white, xlab,
                   ylab, main)}
  
}

# use as ggsurv.svd(kmobject, CI = TRUE)
# will give KM curve with ribbon CI with one legend only; the legend will be as per the geom_step argument.
```

# Code to create a column with the largest /smallest of many columns in dataframe:

for (i in 1:nrow(data)) {
  data$E[i] <- min(data[i,c("A","B","C")])
}




```{r}

# this code is a function which will create a new column E which will contain the smallest of the values of columns A,B,C in dataframe data






#eg.

library(survival)

b <- lung %>% select(age)

bc <- b %>% slice(1:5)

bc$age2 <- c(60,89,33,55,89)

for (i in 1:nrow(bc)) {
  bc$age_min[i] <- min(bc[i,c("age","age2")])
}

```



