---
title: "R script:Cheat-sheet"
date: "`r Sys.Date()`"
author: Salil V Deo MD (svd14 at case dot edu)
output:
  rmdformats::readthedown:
    highlight: kate
    
---


```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")

opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)

opts_knit$set(width=75)

```



This cheat-sheet is a guide to the commonly needed R-code for initial data cleaning and data restructuring. 

## Importing data into Rstudio from various formats:

```{r}
library(tidyverse) # import csv

library(haven) # import SAS, SPSS, STATA 

library(readxl) # import excel files

library(stringr); library(magrittr)

```

Importing syntax:

*csv* -`read_csv("df.csv")`

*SAS* - `read_sas("df.sas7bdat")`

*STATA* - `read_dta("df.dta")`

*SPSS* - `read_sav("df.sav")`

*Excel* - `read_excel("df.xlxs", sheet = 1)`

For excel files, the sheet argument can be either a number of a string with the name of the sheet.

## Converting variable column names to lowercase:

Column names can all be converted to lower or uppercase as needed.
R is case sensitive , hence , columns *year, Year, YEAR* are actually the name of 3 uinque columns in R.


```{r}

df <- mtcars

# to change to UPPER

names(df) <- toupper(names(df))

names(df)

# to change to lower

names(df) <- tolower(names(df))

names(df)

```



```{r}

# to make only first letter CAPITAL for every colname.

# using the stringr library

library(stringr)

names(df) <- stringr::str_to_title(names(df))

names(df)


```

## Look at some dplyr verbs:

### Filter

*Filter* helps to select rows for an argument.

```{r}

# again using mtcars.

summary(df$Mpg)

summary(df$Cyl)

# now to keep only rows with Mpg < 20.

df2 <- df %>% filter(Mpg < 20)

summary(df2$Mpg)

# to keep only Mpg < 20 and Cyl == 6

df3 <- df %>% filter(Mpg < 20) %>%
  filter(Cyl == 6)

summary(df3[, c("Mpg","Cyl")]) # using summary like this can provide the summary of only those columns that you want.

```

### Select:- 

*Select* will keep only those columns in the argument.

```{r}

# using whole name of columns.

df4 <- df %>% select(Mpg, Cyl)

str(df4) # now contains only these two columns.

# now to only select according to type of column.
# am going to convert cylinder to factor.

df$Cyl <- as.factor(df$Cyl)

df4 <- df %>% select_if(is.factor)

str(df4)

# on Cyl included as that one is the only factor in the df.

df4 <- df %>% select_if(is.numeric)

str(df4) # now all but Cyl is included as that is a factor.

# we can also choose based on the name or part of the column name.

df4 <- df %>% select(contains("Mp"))

str(df4) # as only Mpg contains "Mp" that is the only column selected.

df4 <- df %>% select(starts_with("W"))

str(df4) # as Wt is the only column that starts with W.

df4 <- df %>% select(ends_with("c"))

str(df4)

```

### Mutate:-

*Mutate* will help to create new variables. I do not use it as much.
Rather if find $ifelse$ very useful to create variables using arguments based on another variables.

```{r}

summary(df)

# now to create new variable gas_guzzler when Mpg < 20.

df2 <- df %>% mutate(gas_guzzler = ifelse(Mpg < 20, 1, 0))

str(df2) 

# to use mutate, we need to create a new df. Rather than using mutate, I prefer to use base R and ifelse.

df$gas_guzzler <- with(df, ifelse(Mpg < 20, 1, 0))

summary(df$gas_guzzler) # so using this code, I can create new variables without having to keep creating new dataframes. I know some code to create many variables at once, but I like to create 1 variable at a time and code iteratively.


# ifelse arguments can be nested to create as many values as needed.
# this may be the alternative to fct_recode.

df$gas_guzzler2 <- with(df, ifelse(Mpg > 30, 0 , 
  ifelse(Mpg < 20, 1, 2)
)) 


df %>% count(gas_guzzler2) # do, now have this column as factor with 3 levels - 0/1/2

# with this coding, the column is not converted to factor type , but there is really no need to convert it. If some functions need it only as a factor, then we can easily convert it using as.factor()...

```

*as.factor, as.numeric* can be used to change the type of the variable if needed. 

While there are many other dplyr verbs, I have not used them very often and from my experience, are not much needed to regular data cleaning and variable creation.


## Working with Dates:

The *lubridate* package is excellent for working with dates.

```{r, message = FALSE}
library(lubridate)

# get dummy dataset with dates.

df <- read_csv("H:\\R_cheat_sheet\\date_df.csv")

glimpse(df) 

# now first to declare the columns as date formats.

df$date_of_surgery <- as_date(df$date_of_surgery)

df$dob <- as_date(df$dob)

df$last_contact_date <- as_date(df$last_contact_date)
  

# now before we use these columns we can do some simple date 
# function.

today() # this is todays date, or when the rmarkdown was last knitted. 

# Dates in R are presented by convention as ymd.
# ymd format is unambigious as it is conventionally used for date in most software programs.

ymd(20210128) # this can be used to change the numbers to dates.

# to extract the day, month or year from a date.

month("2019-02-10")

year("2011-10-01")

day("2011-10-01")


# this can be used to create new columns...

df$dob_year = year(df$dob)
  
head(df$dob_year, 5)

# we can use month or day to similarly extract the day or month and that can be inserted into a new column.

```

Subtracting dates to get time in days, years.

```{r}

# to subtract dates.

df$age_at_surgery <- (df$dob %--% df$date_of_surgery)/ddays(1)

# now to convert to years, divide by 365.24

df$age_years <- df$age_at_surgery/365.24 # the .24 accounts for the leap years.


# we conventionally present age a whole number.

df$age_years <- round(df$age_years, 0)

summary(df$age_years) # age as whole numbers from the dob and date of surgery.

# if we want to extract months, we can divide by 30.
```
*Lubridate* also has excellent functions to work with hh:mm:ss to calculate difference between date:times. I have not used those much.

```{r}


# to obtain age of a patient today, we can use the today function...

df$cur_age <- (df$dob %--% today())/ddays(1)

df$cur_age <- round((df$cur_age/365.24),0)

summary(df$cur_age) # of course all patients lived, so we have a patient aged 103 years :))

```


